---
title: "Personalized Monitoring of HIV Viral Load Through Prognostic Healthcare (#179)"
author: "IeDEA | FIMP"
output: 
  rmdformats::material:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango

---

```{r setup, include=FALSE}
options(java.parameters = "-Xmx15g")

knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      echo=FALSE,
                      #dpi=96,
                     # fig.width=7,# fig.height=4, # Default figure widths
                     # dev="png", #dev.args=list(type="cairo"), # The png device
                      # Change to dev="postscript" if you want the EPS-files
                      # for submitting. Also remove the dev.args() as the postscript
                      # doesn't accept the type="cairo" argument.
                      error=FALSE)
 
# Evaluate the figure caption after the plot
#knitr::opts_knit$set(eval.after='fig.cap')
 
# Use the table counter that the htmlTable() provides
options(table_counter = TRUE)
 
# Use the figCapNo() with roman letters
#options(fig_caption_no_roman = TRUE)
#options(kableExtra.latex.load_packages = F)

# Then install the Grmd-package by running below code:
#devtools::install_github("gforge/Grmd")


# function to install missing packages
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE, repos='http://cran.rstudio.com/')
  sapply(pkg, require, character.only = TRUE)
}

#install.packages('package_name', dependencies=TRUE, repos='http://cran.rstudio.com/')

packages =c( "dplyr",  "readxl","Hmisc","Gmisc", "magrittr", "flextable", "MASS", "tidyverse", "caret", "knitr", "kableExtra","xtable", "stargazer", "ggpubr", "haven", "PerformanceAnalytics", "naniar", "gridExtra", "ggthemes", "TSstudio")

ipak(packages)

# packages not pubished
#devtools::install_github("kassambara/easyGgplot2")
library(easyGgplot2)

# packages not pubished
#devtools::install_github('cttobin/ggthemr')
#library(ggthemr)


select = dplyr::select; summarize = dplyr::summarize; rename = dplyr::rename; mutate = dplyr::mutate;

#ggthemr("flat")

#source("unbalanced_functions.R")
source("table1_utils.R")
```

# Explaratory Data Analysis

The goal of the exploratory analysis is to identify the key predictors, generate insights and dicover trends and patters. Throughout this process we are mainly interested in detecting:

* Correlation and assoicitations
* Trends in time e.g viral load
* Outliers
* Any other key information that might be useful in ML modeling


```{r warning=FALSE}
rural.cb <- tibble::tribble(
              ~code, ~label,
                     1,'Urban', 2,'Mostly urban', 3,'Mostly rural', 4,'Rural', 9,'Unknown'
              )

level.cb <- tibble::tribble(
              ~code, ~label,
                     1,'Health centre',2,'District hospital',3,'Regional, provincial or university hospital',9,'Unknown'
              )

civstfmt.cb <- tibble::tribble(
              ~code, ~label,
                      1,'Never Married and Not Living w/Partner',2,'Legally Married',3,'Living w/Partner',4,'Separated',5,'Divorced',6,'Widowed'
              )

entptfmt.cb <- tibble::tribble(
              ~code, ~label,
                        1,'outpatient care',2,'inpatient care',3,'pmtct care',4,'TB ward care',
                        5,'community-based org',6,'private care',7,'self-referral',8,'other',
                        9,'RTC (routine testing and counseling)',10,'VCT',11,'HCT(home-based)',
                        12,'MCH (mother child health)',13,'TB clinic',14,'Research Institute',15,'Family Member',
                        16,'Transfer-In',17,'STI Clinic',18,'Unknown',19,'Exposed infant',20,'Under 5 yrs',
                        21,'IDU outreach',22,'YYC(Young Child Clinic)'
              )


pillfmt.cb <- tibble::tribble(
              ~code, ~label,
                      1,'none',2,'very few',3,'about half',4,'most',5,'all'
              )

adhfmt.cb <- tibble::tribble(
              ~code, ~label,
                     1,'Good: 3 or fewer missed doses since last visit',
                     2,'Poor: 4 or more missed doses since last visit',3,'Good: fewer than 2 missed days',
                     4,'Poor: 2 or more missed days'
              )

arvadfmt.cb <- tibble::tribble(
              ~code, ~label,
                     1,'GOOD (>, 95%)',2,'FAIR (85-94%)',3,'POOR (< 85%)'
              )

ynfmt.cb <- tibble::tribble(
              ~code, ~label,
                      1,'Yes',2,'No'
              )

zeroonefmt.cb <- tibble::tribble(
              ~code, ~label,
                      1,'Yes',0,'No'
              )
```

## Assumptions

```{r warning=F}
getFirstXnonNA=function(data,n=3,var){
  #first = dplyr::first(na.omit(value))
  return(data %>%
                  arrange(ptidno, `apptdate`) %>%
                  group_by(ptidno) %>%
                  filter(visit_num <= n) %>%
                  mutate(
                    var_new=dplyr::first(na.omit(!!as.symbol(var)))
                  )%>%
                  ungroup())%>%
                  filter(visit_num == 1)

  #filter(apptdate == min(apptdate)) %>%
}

getFirstNonNAInXMonths=function(data,m=3,var){
  #first = dplyr::first(na.omit(value))
  days=m*30
  return(data %>%
                  arrange(ptidno, `apptdate`) %>%
                  group_by(ptidno) %>%
                  filter(apptdate <= min(apptdate)+days) %>%
                  mutate(
                    var_new=dplyr::first(na.omit(!!as.symbol(var)))
                  )%>%
                  ungroup())%>%
                  filter(visit_num == 1)

  #filter(apptdate == min(apptdate)) %>%
}

getBaselineOnArvStartWithinXMonths=function(data,m=3,var, refdate){
  #first = dplyr::first(na.omit(value))
  days=m*30
  return(data %>%
                  arrange(ptidno, `apptdate`) %>%
                  group_by(ptidno) %>%
                  filter(apptdate <= !!as.symbol(refdate)) %>%
                  mutate(
                    var_new=dplyr::first(na.omit(!!as.symbol(var)))
                  )%>%
                  ungroup())%>%
                  filter(visit_num == 1)

  #filter(apptdate == min(apptdate)) %>%
}



binarize <- function(x) {
  case_when(
               x=='1'~ "Yes",
               x=='0'~ "No",
              is.na(x) ~ "No",
              TRUE ~ x)
  
}


```


```{r warning=F,cache=T}
# Import Dataset
patient.cross.init =haven::read_sas("ak_vlmonitord.sas7bdat")
patient.long.init =haven::read_sas("ak_vlmonitorl.sas7bdat")

```


```{r warning=F}
# codes
patient.long.init$PillsTakenLast7Days <- factor(pillfmt.cb[match(patient.long.init$PillsTakenLast7Days, pillfmt.cb$code), ]$label)
patient.long.init$art_adherence2 <- factor(adhfmt.cb[match(patient.long.init$art_adherence2, adhfmt.cb$code), ]$label)
patient.long.init$arvadherence <- factor(arvadfmt.cb[match(patient.long.init$arvadherence, arvadfmt.cb$code), ]$label)


# merge dataset
patient.long = patient.long.init %>%
                  arrange(ptidno, `apptdate`) %>%
                  group_by(ptidno) %>%
                  mutate(visit_num = row_number())%>%
                  mutate(
                    # vl date
                    vl1_date =  as.Date(dplyr::nth(na.omit(ifelse(is.na(viralload), NA , apptdate)),1), origin="1970-01-01"),
                    vl2_date =  as.Date(dplyr::nth(na.omit(ifelse(is.na(viralload), NA , apptdate)),2), origin="1970-01-01"),
                    vl3_date =  as.Date(dplyr::nth(na.omit(ifelse(is.na(viralload), NA , apptdate)),3), origin="1970-01-01"),
                    
                    # num days to vl
                    
                    num_days_to_vl1 =  dplyr::nth(na.omit(ifelse(is.na(viralload), NA , apptdate-min(apptdate))),1),
                    num_days_to_vl2 =  dplyr::nth(na.omit(ifelse(is.na(viralload), NA , apptdate-min(apptdate))),2),
                    num_days_to_vl3 =  dplyr::nth(na.omit(ifelse(is.na(viralload), NA , apptdate-min(apptdate))),3),
                    
                    # difference in days between vl2 and vl2
                    days_from_vl1_vl2 = num_days_to_vl2- num_days_to_vl1,
                    days_from_vl2_vl3 = num_days_to_vl3- num_days_to_vl2,
      
                    
                  
                    # vl count
                    vl_count1 = dplyr::nth(na.omit(viralload),1),
                    vl_count2 = dplyr::nth(na.omit(viralload),2),
                    vl_count3 = dplyr::nth(na.omit(viralload),3),
                    
                    vl_failure1 = factor(ifelse(vl_count1<=1000, "No","Yes")),
                    vl_failure2 = factor(ifelse(vl_count2<=1000, "No","Yes")),
                    vl_failure3 = factor(ifelse(vl_count3<=1000, "No","Yes")),
                    
                    # num encunters to vl
                    num_encs_to_vl1 = dplyr::nth(na.omit(ifelse(is.na(viralload), NA , visit_num-1)),1),
                    num_encs_to_vl2 = dplyr::nth(na.omit(ifelse(is.na(viralload), NA , visit_num-1)),2),
                    num_encs_to_vl3 = dplyr::nth(na.omit(ifelse(is.na(viralload), NA , visit_num-1)),3),
                    
                    # factor
                    WHOstage= factor(WHOstage)
                    
                    
                  )%>%
                  ungroup() %>% left_join(patient.cross.init%>%select(ptidno, arvstart),  c("ptidno"="ptidno"))
                  
                  
# handle AMPATH's data missing xrossection civilstatus
civilstatus.cx=getFirstNonNAInXMonths(patient.long,3,'civilstatus')%>%select(ptidno, civilstatus=var_new)


# handle all sites missing baseline features
Height.cx=getFirstNonNAInXMonths(patient.long,3,'Height')%>%select(ptidno, Height=var_new)
Weight.cx=getFirstNonNAInXMonths(patient.long,3,'Weight')%>%select(ptidno, Weight=var_new)
bmi.cx=getFirstNonNAInXMonths(patient.long,3,'bmi')%>%select(ptidno, bmi=var_new)
nhif.cx=getFirstNonNAInXMonths(patient.long,3,'nhif')%>%select(ptidno, nhif=var_new)



# perform joining on the missing varibles
patient.cross = patient.cross.init%>%
                   mutate(transfer=ifelse(is.na(transfer)==T,0,1))%>%
                   left_join(civilstatus.cx,  c("ptidno"="ptidno"))%>%
                   left_join(Height.cx,  c("ptidno"="ptidno"))%>%
                   left_join(Weight.cx,  c("ptidno"="ptidno"))%>%
                   left_join(bmi.cx,  c("ptidno"="ptidno"))%>%
                   left_join(nhif.cx,  c("ptidno"="ptidno"))%>%
  
                   mutate(
                      civilstatusENROL=ifelse(program=="AMPATH",civilstatus,civilstatusENROL)
                     
                    )



#merge master dataset
dataset=patient.cross %>%  left_join(select(filter(patient.long, visit_num == 1),
                              num_days_to_vl1,num_days_to_vl2,num_days_to_vl3,vl_failure1,
                              vl_failure2,vl_failure3,
                              days_from_vl1_vl2,days_from_vl2_vl3,
                              vl_count1,vl_count2,vl_count3,ptidno,vl1_date,vl2_date,
                              vl3_date,
                              num_encs_to_vl1,num_encs_to_vl2,num_encs_to_vl3
  
                            ),  c("ptidno"="ptidno"))
#Replace numerical codes with strings from codebook
dataset$civilstatusENROL <- factor(civstfmt.cb[match(dataset$civilstatusENROL, civstfmt.cb$code), ]$label)
dataset$male <- factor(zeroonefmt.cb[match(dataset$male, zeroonefmt.cb$code), ]$label)
dataset$transfer <- factor(zeroonefmt.cb[match(dataset$transfer, zeroonefmt.cb$code), ]$label)
dataset$Death <- factor(zeroonefmt.cb[match(dataset$Death, zeroonefmt.cb$code), ]$label)
dataset$nhif <- factor(zeroonefmt.cb[match(dataset$nhif, zeroonefmt.cb$code), ]$label)
dataset$entrypoint = factor(entptfmt.cb[match(dataset$entrypoint, entptfmt.cb$code), ]$label) 

dataset= dataset %>% 
                 mutate(across(c('Death','male', 'entrypoint', "civilstatusENROL", "transfer", "nhif"), as.factor)) 
                  # assuming if not 1 is 0

# Relevel
varList= list(
"Yes"=c('male','transfer','Death')
)
#dataset= relevelBy(varList,dataset) 


patient.long.cross = patient.long %>%  left_join(patient.cross, c("ptidno"="ptidno"))
#str(dataset)
```




* AMPATH civil status is collected longitudinal; as such missing data on the crossectional dataset has to be replaced with longitudinal civil status recorded not > 3 months after the first encounter
* We have assumed if is.na(transfer) then No else Yes
* Unrecorded data (occurring at the 1st encounter) on Height, Weight, BMI, NHIF have to be replaced with non-null next encounter recorded < 3 months from the first encounter
* Viral Load index  - It is not possible to know when the next VL is drawn longitudinally, as shown in the table below. As such, we are treating each VL result as individual tests ordered independently.

```{r warning=F, eval=T}
patient.long %>% select(ptidno,visit_num,apptdate,VLundetectable,viralload,viralloadc,Height)%>%  
   filter(ptidno==52764) %>% arrange(ptidno, `apptdate`)  %>% kable () %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

# Viral Load Counts

We will start by looking at the overall distribution of VL

## Distribution


```{r fig.align='center',warning=FALSE, cache=T}
hiv_dfl=dataset%>%filter(!is.na(vl_count1))


a =hiv_dfl%>% mutate(VL_Failure=ifelse(vl_count1<=1000, "No","Yes")) %>%
  ggboxplot( palette = "jco",  ylab = "VL Count",
          y = c("vl_count1"),  
          x = c("VL_Failure"), 
          shape = "VL_Failure",	
          yscale = c("log2"),
          add = "jitter", 
          color = "VL_Failure",  add.params = list(size = 0.2, alpha=0.3, jitter = 0.2) )+
  scale_color_manual(values= c("#00AFBB", "#E7B800"))+ scale_fill_manual(values= c("#00AFBB", "#E7B800"))


b =hiv_dfl%>% mutate(VL_Failure=ifelse(vl_count2<=1000, "No","Yes")) %>%
  ggboxplot( palette = "jco",  ylab = "VL Count 2",
          y = c("vl_count2"),  
          x = c("VL_Failure"), 
          shape = "VL_Failure",	
          yscale = c("log2"),
          add = "jitter", 
          color = "VL_Failure",  add.params = list(size = 0.2, alpha=0.3, jitter = 0.2) )	+
scale_color_manual(values= c("#00AFBB", "#E7B800"))+ scale_fill_manual(values= c("#00AFBB", "#E7B800"))


c = hiv_dfl%>% mutate(VL_Failure=ifelse(vl_count1<=1000, "No","Yes"))  %>%
  ggplot2.histogram( xName="vl_count1", backgroundColor="white",  bins=80,
                         brewerPalette="Paired",
                         groupName="VL_Failure", xScale="log2",rug = TRUE ,    
                         alpha=0.5, addDensity=TRUE,  removePanelGrid=TRUE ,removePanelBorder=TRUE,
                         addMeanLine=TRUE, meanLineSize=1) + xlab("VL Count 1")+
  scale_color_manual(values= c("#00AFBB", "#E7B800")) + scale_fill_manual(values= c("#00AFBB", "#E7B800"))

d = hiv_dfl%>% mutate(VL_Failure=ifelse(vl_count2<=1000, "No","Yes"))  %>%
  ggplot2.histogram( xName="vl_count2", backgroundColor="white",  bins=80,
                         brewerPalette="Paired",
                         groupName="VL_Failure", xScale="log2", xlab = "VL Count 2",rug = TRUE ,
                         alpha=0.5, addDensity=TRUE,  removePanelGrid=TRUE ,removePanelBorder=TRUE,
                         addMeanLine=TRUE, meanLineSize=1) + xlab("VL Count 2")+
  scale_color_manual(values= c("#00AFBB", "#E7B800"))+ scale_fill_manual(values= c("#00AFBB", "#E7B800"))

ggarrange(a,b,c,d,
             common.legend = TRUE, legend="bottom",	
             ncol = 2, nrow=2)	

```

Is a cutoff of 1000 valid with relation to VL detectability limit? To understll start by decomposing the viral load counts into its three components that is inline with the goal of our analyis- Viral load 1, viral load 2 and viral load 3.


## Overall Viral Load Counts Detection Limit

```{r fig.align='center',warning=FALSE, cache=T}

patient.long.cross %>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 500 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()


```

## MBARARA Viral Load Detection Limit

```{r fig.align='center',warning=FALSE, cache=T}

 patient.long.cross %>%
  filter(program=="MBARARA")%>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 600 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()

```

## MASAKA Viral Load Detection Limit

```{r fig.align='center',warning=FALSE,cache=T}

patient.long.cross %>%
  filter(program=="MASAKA")%>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 70 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()


```

## RAKAI Viral Load Detection Limit

```{r fig.align='center',warning=FALSE,cache=T}


a = patient.long.cross %>%
  filter(program=="RAKAI")%>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 1200 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()

b = patient.long.cross %>%
  filter(program=="RAKAI")%>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 80 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()

ggarrange(a,b,
             common.legend = TRUE, 
             ncol = 2, nrow=1)	

```

## IDI Viral Load  Detection Limit

```{r fig.align='center',warning=FALSE,cache=T}

patient.long.cross %>%
  filter(program=="IDI")%>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 100 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()

```


## AMPATH Viral Load Detection Limit

```{r fig.align='center',warning=FALSE,cache=T}


a = patient.long.cross %>%
  filter(program=="AMPATH")%>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 20 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()

b = patient.long.cross %>%
  filter(program=="AMPATH")%>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 50 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()

ggarrange(a,b,
             common.legend = TRUE, 
             ncol = 2, nrow=1)	

```



## FACES Viral Load Detection Limit

```{r fig.align='center',warning=FALSE,cache=T}

patient.long.cross %>%
  filter(program=="FACES")%>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 70 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()


```




## KISESA Viral Load Detection Limit

```{r fig.align='center',warning=FALSE,cache=T}

patient.long.cross %>%
  filter(program=="KISESA")%>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 60 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()


```

## MOROGORO Viral Load Detection Limit

```{r fig.align='center',warning=FALSE,cache=T}

patient.long.cross %>%
  filter(program=="MOROGORO")%>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 60 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()


```

## TUMBI Viral Load Detection Limit

```{r fig.align='center',warning=FALSE,cache=T}

patient.long.cross %>%
  filter(program=="TUMBI")%>%
  filter(!is.na(viralload))%>%
  mutate(year = as.numeric(format(apptdate,'%Y')), onARV=factor(ifelse(onARV==1, "Yes", "No")))%>% 
  filter(!is.na(viralload))%>%
  filter( viralload <= 60 ) %>% 
    ggplot(aes(x=apptdate, y=viralload)) + 
    geom_jitter(alpha=0.3, aes(color=onARV)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()

```


We also need to look if this is consisten with VL index

## Viral Load 1 Detection Limit

```{r fig.align='center',warning=FALSE, cache=T}

dataset %>%
  mutate(year = as.numeric(format(vl1_date,'%Y')), on_ART=factor(ifelse(arvstartTHERAP<=vl1_date, "Yes", "No")))%>% 
  filter(!is.na(vl_count1))%>%
  filter( vl_count1 <= 500 ) %>% 
    ggplot(aes(x=vl1_date, y=vl_count1)) + 
    geom_jitter(alpha=0.3, aes(color=on_ART)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()


```

## Viral Load 2 Detection Limit

```{r fig.align='center',warning=FALSE, cache=T}

dataset %>%
  mutate(year = as.numeric(format(vl2_date,'%Y')), on_ART=factor(ifelse(arvstartTHERAP<=vl2_date, "Yes", "No")))%>% 
  filter(!is.na(vl_count2))%>%
  filter( vl_count2 <= 500 ) %>% 
    ggplot(aes(x=vl2_date, y=vl_count2)) + 
    geom_jitter(alpha=0.3, aes(color=on_ART)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()


```


## Viral Load 3 Detection Limit

```{r fig.align='center',warning=FALSE, cache=T}

dataset %>%
  mutate(year = as.numeric(format(vl3_date,'%Y')), on_ART=factor(ifelse(arvstartTHERAP<=vl3_date, "Yes", "No")))%>% 
  filter(!is.na(vl_count3))%>%
  filter( vl_count3 <= 500 ) %>% 
    ggplot(aes(x=vl3_date, y=vl_count3)) + 
    geom_jitter(alpha=0.3, aes(color=on_ART)) +
    geom_smooth() +
    facet_wrap( ~ program, scales = "free_x")+
      theme_clean()


```



## Viral Load Count | Programs

```{r fig.align='center',warning=FALSE}
na.rm.var=F;stats=F

# Getting descriptive statistics 
mergeDesc(
  
  "First VL count" = getTable1Stats("vl_count1","program",desc_both,statistics=stats,na.rm=na.rm.var),
  "Second VL count" = getTable1Stats("vl_count2","program",desc_both,statistics=stats,na.rm=na.rm.var),
  "Third VL count" = getTable1Stats("vl_count3","program",desc_both,statistics=stats,na.rm=na.rm.var),
  "Days (from enrollment) to first VL" = getTable1Stats("num_days_to_vl1","program",desc_both,statistics=stats,na.rm=na.rm.var),       
  "Days (from enrollment) to second VL" = getTable1Stats("num_days_to_vl2","program",desc_both,statistics=stats,na.rm=na.rm.var),   
  "Days (from enrollment) to third VL" = getTable1Stats("num_days_to_vl3","program",desc_both,statistics=stats,na.rm=na.rm.var), 
  "Days (from 1st VL) to second VL" = getTable1Stats("days_from_vl1_vl2","program",desc_both,statistics=stats,na.rm=na.rm.var),
  "Days (from 2nd VL) to thrid VL" = getTable1Stats("days_from_vl2_vl3","program",desc_both,statistics=stats,na.rm=na.rm.var),
   "# of encounter B4 first VL" = getTable1Stats("num_encs_to_vl1","program",desc_both,statistics=stats,na.rm=na.rm.var),
  "# of encounter B4 second VL" = getTable1Stats("num_encs_to_vl2","program",desc_both,statistics=stats,na.rm=na.rm.var),
  "# of encounter B4 third  VL" = getTable1Stats("num_encs_to_vl3","program",desc_both,statistics=stats,na.rm=na.rm.var)
  #"NHIF at enrollment" = getTable1Stats("nhif","program",desc_both,statistics=stats,na.rm=na.rm.var)
          
          )%>%
   htmlTable( caption  = "<b> Socio-demographics </b>",useViewer=T,ctable = TRUE, align = 'lcccc',
              n.cgroup = c(1,9), cgroup = c('', 'Programs') ,tfoot="<sup>&Dagger;</sup>"
              )%>% htmlTable::addHtmlTableStyle(col.rgroup = c("#FFF", "#EEF")) 



```



# Missing Data

## Socio-demographics Missing Data Summary

```{r fig.align='center',warning=FALSE, cache=T}
x=c('male','Death','civilstatusENROL', 'transfer','Height','Weight', 'entrypoint', 'nhif', 'bmi','program', 'ageatarvstart' )
y=c('vl_count1','vl_count2','vl_count3')
kable( dataset%>%select(c(x,y)) %>%naniar::miss_var_summary()%>%top_n(20,n_miss))

naniar::gg_miss_var(dataset%>%select(c(x,y))  )+ labs(title = "Frequency of Missingness")
naniar::gg_miss_fct(x = dataset%>%select(c(x,y, "program")), fct = program)

p1 = naniar::gg_miss_fct(x = dataset%>%select(c(x,y, "vl_failure1")), fct = vl_failure1)

p2 =naniar::gg_miss_fct(x = dataset%>%select(c(x,y, "vl_failure2")), fct = vl_failure2)

p3 =naniar::gg_miss_fct(x = dataset%>%select(c(x,y, "vl_failure3")), fct = vl_failure3)
p4 =naniar::gg_miss_fct(x = dataset%>%select(c(x,y, "Death")), fct = Death)



gridExtra::grid.arrange(p1,p2,p3,
             p4 ,
             ncol = 2, nrow=2)
```




# Bivariate Analysis

To understand the relationship of each variable with the outcome variable:


## Socio-demographics

### Overall Socio-demographics Correlation

```{r fig.align='center',warning=FALSE, cache=T}
PerformanceAnalytics::chart.Correlation(dataset%>%
                                          mutate(
                                            male=ifelse(male=="Yes", 1,0),
                                            Death=ifelse(Death=="Yes", 1,0),
                                            transfer=ifelse(transfer=="Yes", 1,0),
                                            married=ifelse(civilstatusENROL=="Legally Married", 1,0)
                                          )%>%
                                          select(Weight,bmi,Height,male,Death,married,"ageatarvstart", vl_count2, vl_count1), histogram=TRUE, pch=22)	



```

### Socio-demographics | Viral Load Failure 1
```{r fig.align='center',warning=FALSE, cache=T}
na.rm.var=F;stats=T

# Getting descriptive statistics 
mergeDesc(
  
  "age at first exposure to ARV" = getTable1Stats("ageatarvstart","vl_failure1",desc_both,statistics=stats,na.rm=na.rm.var),  
  "male gender" = getTable1Stats("male","vl_failure1",desc_both,statistics=stats,na.rm=na.rm.var),       
  "death" = getTable1Stats("Death","vl_failure1",desc_both,statistics=stats,na.rm=na.rm.var),   
  "civil status at enrollment" = getTable1Stats("civilstatusENROL","vl_failure1",desc_both,statistics=stats,na.rm=na.rm.var), 
  "patient transferred out" = getTable1Stats("transfer","vl_failure1",desc_both,statistics=stats,na.rm=na.rm.var),
  "Height (cm) at enrollment" = getTable1Stats("Height","vl_failure1",desc_both,statistics=stats,na.rm=na.rm.var),
  "Weight (kg) at enrollment" = getTable1Stats("Weight","vl_failure1",desc_both,statistics=stats,na.rm=na.rm.var),
  "BMI at enrollment" = getTable1Stats("bmi","vl_failure1",desc_both,statistics=stats,na.rm=na.rm.var),
   #"First VL count" = getTable1Stats("vl_count1","program",desc_both,statistics=stats,na.rm=na.rm.var),
  "Second VL count" = getTable1Stats("vl_count2","vl_failure1",desc_both,statistics=stats,na.rm=na.rm.var),
  "Third VL count" = getTable1Stats("vl_count3","vl_failure1",desc_both,statistics=stats,na.rm=na.rm.var)
  #"entry point to care" = getTable1Stats("entrypoint","program",desc_both,statistics=stats,na.rm=na.rm.var)
  #"NHIF at enrollment" = getTable1Stats("nhif","program",desc_both,statistics=stats,na.rm=na.rm.var)
          
          )%>%
   htmlTable( caption  = "<b> Socio-demographics </b>",useViewer=T,ctable = TRUE, align = 'lcccc',
              n.cgroup = c(1,2,1 ), cgroup = c('', 'First Virologic Failure', '') ,tfoot="<sup>&Dagger;</sup>"
              )%>% htmlTable::addHtmlTableStyle(col.rgroup = c("#FFF", "#EEF")) 



```


```{r fig.align='center',warning=FALSE}

naniar::gg_miss_upset(dataset%>%select(x, 'vl_failure1'), nsets = 5,nintersects = NA)



```

### Socio-demographics | Viral Load Failure 2
```{r fig.align='center',warning=FALSE}
na.rm.var=F;stats=T

# Getting descriptive statistics 
mergeDesc(
  "age at first exposure to ARV" = getTable1Stats("ageatarvstart","vl_failure2",desc_both,statistics=stats,na.rm=na.rm.var),
  "male gender" = getTable1Stats("male","vl_failure2",desc_both,statistics=stats,na.rm=na.rm.var),       
  "death" = getTable1Stats("Death","vl_failure2",desc_both,statistics=stats,na.rm=na.rm.var),   
  "civil status at enrollment" = getTable1Stats("civilstatusENROL","vl_failure2",desc_both,statistics=stats,na.rm=na.rm.var), 
  "patient transferred out" = getTable1Stats("transfer","vl_failure2",desc_both,statistics=stats,na.rm=na.rm.var),
  "Height (cm) at enrollment" = getTable1Stats("Height","vl_failure2",desc_both,statistics=stats,na.rm=na.rm.var),
  "Weight (kg) at enrollment" = getTable1Stats("Weight","vl_failure2",desc_both,statistics=stats,na.rm=na.rm.var),
  "BMI at enrollment" = getTable1Stats("bmi","vl_failure2",desc_both,statistics=stats,na.rm=na.rm.var),
   "First VL count" = getTable1Stats("vl_count1","vl_failure2",desc_both,statistics=stats,na.rm=na.rm.var),
  #"Second VL count" = getTable1Stats("vl_count2","program",desc_both,statistics=stats,na.rm=na.rm.var),
  "Third VL count" = getTable1Stats("vl_count3","vl_failure2",desc_both,statistics=stats,na.rm=na.rm.var)
  #"entry point to care" = getTable1Stats("entrypoint","program",desc_both,statistics=stats,na.rm=na.rm.var)
  #"NHIF at enrollment" = getTable1Stats("nhif","program",desc_both,statistics=stats,na.rm=na.rm.var)
          
          )%>%
   htmlTable( caption  = "<b> Socio-demographics </b>",useViewer=T,ctable = TRUE, align = 'lcccc',
              n.cgroup = c(1,2,1 ), cgroup = c('', 'Second Virologic Failure', '') ,tfoot="<sup>&Dagger;</sup>"
              )%>% htmlTable::addHtmlTableStyle(col.rgroup = c("#FFF", "#EEF")) 



```




```{r fig.align='center',warning=FALSE}

naniar::gg_miss_upset(dataset%>%select(x, 'vl_failure2'), nsets = 5,nintersects = NA)

```

### Socio-demographics | Viral Load Failure 3
```{r fig.align='center',warning=FALSE}
na.rm.var=F;stats=T

# Getting descriptive statistics 
mergeDesc(
  "age at first exposure to ARV" = getTable1Stats("ageatarvstart","vl_failure3",desc_both,statistics=stats,na.rm=na.rm.var),
  "male gender" = getTable1Stats("male","vl_failure3",desc_both,statistics=stats,na.rm=na.rm.var),       
  "death" = getTable1Stats("Death","vl_failure3",desc_both,statistics=stats,na.rm=na.rm.var),   
  "civil status at enrollment" = getTable1Stats("civilstatusENROL","vl_failure3",desc_both,statistics=stats,na.rm=na.rm.var), 
  "patient transferred out" = getTable1Stats("transfer","vl_failure3",desc_both,statistics=stats,na.rm=na.rm.var),
  "Height (cm) at enrollment" = getTable1Stats("Height","vl_failure3",desc_both,statistics=stats,na.rm=na.rm.var),
  "Weight (kg) at enrollment" = getTable1Stats("Weight","vl_failure3",desc_both,statistics=stats,na.rm=na.rm.var),
  "BMI at enrollment" = getTable1Stats("bmi","vl_failure3",desc_both,statistics=stats,na.rm=na.rm.var),
  "First VL count" = getTable1Stats("vl_count1","vl_failure3",desc_both,statistics=stats,na.rm=na.rm.var),
  "Second VL count" = getTable1Stats("vl_count2","vl_failure3",desc_both,statistics=stats,na.rm=na.rm.var)
  #"Third VL count" = getTable1Stats("vl_count3","program",desc_both,statistics=stats,na.rm=na.rm.var)
  #"entry point to care" = getTable1Stats("entrypoint","program",desc_both,statistics=stats,na.rm=na.rm.var)
  #"NHIF at enrollment" = getTable1Stats("nhif","program",desc_both,statistics=stats,na.rm=na.rm.var)
          
          )%>%
   htmlTable( caption  = "<b> Socio-demographics </b>",useViewer=T,ctable = TRUE, align = 'lcccc',
              n.cgroup = c(1,2,1 ), cgroup = c('', 'Third Virologic Failure', '') ,tfoot="<sup>&Dagger;</sup>"
              )%>% htmlTable::addHtmlTableStyle(col.rgroup = c("#FFF", "#EEF")) 



```




```{r fig.align='center',warning=FALSE}

naniar::gg_miss_upset(dataset%>%select(x, 'vl_failure3'), nsets = 5,nintersects = NA)

```

### Socio-demographics | Program
```{r fig.align='center',warning=FALSE}
na.rm.var=F;stats=F

# Getting descriptive statistics 
mergeDesc(
  "age at first exposure to ARV" = getTable1Stats("ageatarvstart","program",desc_both,statistics=stats,na.rm=na.rm.var),
  "male gender" = getTable1Stats("male","program",desc_both,statistics=stats,na.rm=na.rm.var),       
  "death" = getTable1Stats("Death","program",desc_both,statistics=stats,na.rm=na.rm.var),   
  "civil status at enrollment" = getTable1Stats("civilstatusENROL","program",desc_both,statistics=stats,na.rm=na.rm.var), 
  "patient transferred out" = getTable1Stats("transfer","program",desc_both,statistics=stats,na.rm=na.rm.var),
  #"Height (cm) at enrollment" = getTable1Stats("'Height","program",desc_both,statistics=stats,na.rm=na.rm.var),
  "Weight (kg) at enrollment" = getTable1Stats("Weight","program",desc_both,statistics=stats,na.rm=na.rm.var),
  "BMI at enrollment" = getTable1Stats("bmi","program",desc_both,statistics=stats,na.rm=na.rm.var)
  #"entry point to care" = getTable1Stats("entrypoint","program",desc_both,statistics=stats,na.rm=na.rm.var)
  #"NHIF at enrollment" = getTable1Stats("nhif","program",desc_both,statistics=stats,na.rm=na.rm.var)
          
          )%>%
   htmlTable( caption  = "<b> Socio-demographics </b>",useViewer=T,ctable = TRUE, align = 'lcccc',
              n.cgroup = c(1,9 ), cgroup = c('', 'Programs') ,tfoot="<sup>&Dagger;</sup>"
              )%>% htmlTable::addHtmlTableStyle(col.rgroup = c("#FFF", "#EEF")) 



```



```{r fig.align='center',warning=FALSE}

naniar::gg_miss_upset(dataset%>%select(x, 'program'), nsets = 5,nintersects = NA)



```




## Clinical Profile

### Overall Clinical Profile Correlation

```{r fig.align='center',warning=FALSE, cache=F}

# binarize response
binary.var =c(
  'Cough','KSdiag','NephroCardiomyopathy','TBsignNone','cervicalcadiag','chronicisosporiasisdiag',
  'coughTB','cryptococcdiag','cryptospordiag','cytomegdiag','encephaldiag','esocandiddiag','ethambutol',
  'ethambutolinh','herpessvdiag','WeightLoss','inh','inhtbtx','lymphomadiag',
  'mdrTbTx','mycobacdiag','mycosisdiag','nightsweats','nightsweatsTB','lymphomadiag','oiProph','on_ARV','otherTbTx',
  'pcpdiag','pmldiag','pyrazinamide','rifabutin','rifafour','rifampicin','rifater','rifinah','rsbPneumoniaDiag','salmonelladiag',
  'streptomycin','tbextrapulmdiag','tbpulmdiag','tbtx','three_FDC','wastingsyndiag','toxoplasdiag','weightlossTB'
)

```

### Clinical Profile Prior to ART Initiation | Viral Load Failure 1
```{r fig.align='center',warning=FALSE, cache=T}

refdate='arvstart'
              
# Generate baseline clinical covariates
Cough.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'Cough', refdate)%>%select(ptidno, Cough=var_new)
KSdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'KSdiag', refdate)%>%select(ptidno, KSdiag=var_new)
NephroCardiomyopathy.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'NephroCardiomyopathy', refdate)%>%select(ptidno, NephroCardiomyopathy=var_new)
PillsTakenLast7Days.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'PillsTakenLast7Days', refdate)%>%select(ptidno, PillsTakenLast7Days=var_new)
TBsignNone.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'TBsignNone', refdate)%>%select(ptidno, TBsignNone=var_new)
WHOstage.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'WHOstage', refdate)%>%select(ptidno, WHOstage=var_new)

cd4.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'cd4', refdate)%>%select(ptidno, cd4=var_new)
cervicalcadiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'cervicalcadiag', refdate)%>%select(ptidno, cervicalcadiag=var_new)
chronicisosporiasisdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'chronicisosporiasisdiag', refdate)%>%select(ptidno, chronicisosporiasisdiag=var_new)
coughTB.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'coughTB', refdate)%>%select(ptidno, coughTB=var_new)

cryptococcdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'cryptococcdiag', refdate)%>%select(ptidno, cryptococcdiag=var_new)
cryptospordiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'cryptospordiag', refdate)%>%select(ptidno, cryptospordiag=var_new)
cytomegdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'cytomegdiag', refdate)%>%select(ptidno, cytomegdiag=var_new)
encephaldiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'encephaldiag', refdate)%>%select(ptidno, encephaldiag=var_new)
esocandiddiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'esocandiddiag', refdate)%>%select(ptidno, esocandiddiag=var_new)
ethambutol.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'ethambutol', refdate)%>%select(ptidno, ethambutol=var_new)
ethambutolinh.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'ethambutolinh', refdate)%>%select(ptidno, ethambutolinh=var_new)
herpessvdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'herpessvdiag', refdate)%>%select(ptidno, herpessvdiag=var_new)

#// new
Weight.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'Weight', refdate)%>%select(ptidno, weight=var_new)
WeightLoss.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'WeightLoss', refdate)%>%select(ptidno, WeightLoss=var_new)
art_adherence.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'art_adherence', refdate)%>%select(ptidno, art_adherence=var_new)
art_adherence2.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'art_adherence2', refdate)%>%select(ptidno, art_adherence2=var_new)
arvadherence.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'arvadherence', refdate)%>%select(ptidno, arvadherence=var_new)
bmi.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'bmi', refdate)%>%select(ptidno, BMI=var_new)
hgb.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'hgb', refdate)%>%select(ptidno, hgb=var_new)
inh.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'inh', refdate)%>%select(ptidno, inh=var_new)
inhtbtx.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'inhtbtx', refdate)%>%select(ptidno, inhtbtx=var_new)
lymphomadiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'lymphomadiag', refdate)%>%select(ptidno, lymphomadiag=var_new)
mdrTbTx.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'mdrTbTx', refdate)%>%select(ptidno, mdrTbTx=var_new)
mycobacdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'mycobacdiag', refdate)%>%select(ptidno, mycobacdiag=var_new)
mycosisdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'mycosisdiag', refdate)%>%select(ptidno, mycosisdiag=var_new)
nightsweats.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'nightsweats', refdate)%>%select(ptidno, nightsweats=var_new)
nightsweatsTB.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'nightsweatsTB', refdate)%>%select(ptidno, nightsweatsTB=var_new)

numARVs.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'numARVs', refdate)%>%select(ptidno, numARVs=var_new)
oiProph.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'oiProph', refdate)%>%select(ptidno, oiProph=var_new)
onARV.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'onARV', refdate)%>%select(ptidno, on_ARV=var_new)
otherTbTx.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'otherTbTx', refdate)%>%select(ptidno, otherTbTx=var_new)
pcpdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'pcpdiag', refdate)%>%select(ptidno, pcpdiag=var_new)
pmldiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'pmldiag', refdate)%>%select(ptidno, pmldiag=var_new)
pyrazinamide.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'pyrazinamide', refdate)%>%select(ptidno, pyrazinamide=var_new)

regimen.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'regimen', refdate)%>%select(ptidno, regimen=var_new)
rifabutin.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rifabutin', refdate)%>%select(ptidno, rifabutin=var_new)
rifafour.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rifafour', refdate)%>%select(ptidno, rifafour=var_new)
rifampicin.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rifampicin', refdate)%>%select(ptidno, rifampicin=var_new)
rifater.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rifater', refdate)%>%select(ptidno, rifater=var_new)
rifinah.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rifinah', refdate)%>%select(ptidno, rifinah=var_new)
rsbPneumoniaDiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rsbPneumoniaDiag', refdate)%>%select(ptidno, rsbPneumoniaDiag=var_new)
salmonelladiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'salmonelladiag', refdate)%>%select(ptidno, salmonelladiag=var_new)

streptomycin.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'streptomycin', refdate)%>%select(ptidno, streptomycin=var_new)
tbextrapulmdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'tbextrapulmdiag', refdate)%>%select(ptidno, tbextrapulmdiag=var_new)
tbpulmdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'tbpulmdiag', refdate)%>%select(ptidno, tbpulmdiag=var_new)
tbtx.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'tbtx', refdate)%>%select(ptidno, tbtx=var_new)
three_FDC.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'three_FDC', refdate)%>%select(ptidno, three_FDC=var_new)
wastingsyndiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'wastingsyndiag', refdate)%>%select(ptidno, wastingsyndiag=var_new)
toxoplasdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'toxoplasdiag', refdate)%>%select(ptidno, toxoplasdiag=var_new)
weightlossTB.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'weightlossTB', refdate)%>%select(ptidno, weightlossTB=var_new)
              

# add some mpee
dataset.cx.arv=dataset%>%
                   left_join(Cough.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(KSdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(NephroCardiomyopathy.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(PillsTakenLast7Days.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(TBsignNone.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(WHOstage.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(cd4.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(cervicalcadiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(chronicisosporiasisdiag.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(coughTB.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(cryptococcdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(cryptospordiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(cytomegdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(encephaldiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(esocandiddiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(ethambutol.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(ethambutolinh.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(herpessvdiag.cx.arv,  c("ptidno"="ptidno"))%>%
  
                    # new
                   left_join(Weight.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(WeightLoss.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(art_adherence.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(art_adherence2.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(arvadherence.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(bmi.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(hgb.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(inh.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(inhtbtx.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(lymphomadiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(mdrTbTx.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(mycobacdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(mycosisdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(onARV.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(nightsweats.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(nightsweatsTB.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(numARVs.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(oiProph.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(otherTbTx.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(pcpdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(pmldiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(pyrazinamide.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(regimen.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rifabutin.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rifampicin.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rifafour.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rifater.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rifinah.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rsbPneumoniaDiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(streptomycin.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(salmonelladiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(tbextrapulmdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(tbpulmdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(tbtx.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(three_FDC.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(wastingsyndiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(toxoplasdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(weightlossTB.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   mutate(
                      civilstatusENROL=ifelse(program=="AMPATH",civilstatus,civilstatusENROL)
                     
                    )%>%
                   mutate(across(binary.var, as.character))%>%
                   mutate(across(binary.var, binarize))%>%
                   mutate(across(binary.var, as.factor))
  
na.rm.var=F;stats=T;dataset.cx=dataset.cx.arv;

# Getting descriptive statistics 
mergeDesc(
  
  # vitals
  #"current ARV status" = getTable1Stats("on_ARV","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "weight" = getTable1Stats("weight","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "Weight Loss" = getTable1Stats("WeightLoss","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),       
  "BMI" = getTable1Stats("BMI","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "hemoglobin" = getTable1Stats("hgb","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "CD4" = getTable1Stats("cd4","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Cough" = getTable1Stats("Cough","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "Night Sweats" = getTable1Stats("nightsweats","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  
  # staging
  "WHO stage" = getTable1Stats("WHOstage","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Staging: Symptomatic HIV-associated Nephropathy/Cardiomyopathy" = getTable1Stats("NephroCardiomyopathy","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  
   # TB symptoms
  "TB sign: none" = getTable1Stats("TBsignNone","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "TB sign: cough" = getTable1Stats("coughTB","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "TB sign: weight loss" = getTable1Stats("weightlossTB","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),

  
   # Diagnosis
  "Diagnosis: recurrent severe bacterial pneumonia" = getTable1Stats("rsbPneumoniaDiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Recurrent Septicemia including non-typhoidal Salmonella" = getTable1Stats("salmonelladiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: HIV Wasting Syndrome" = getTable1Stats("wastingsyndiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Toxoplasmosis, CNS" = getTable1Stats("toxoplasdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "Diagnosis: Extrapulmonary TB" = getTable1Stats("tbextrapulmdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "Diagnosis: Pulmonary TB" = getTable1Stats("tbpulmdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Lymphoma (cerebral or B-cell non-Hodgkin)" = getTable1Stats("lymphomadiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),   
  "Diagnosis: Kaposi's Sarcoma" = getTable1Stats("KSdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),       
  "Diagnosis: Cervical Cancer" = getTable1Stats("cervicalcadiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  #"Diagnosis: Chronic Isosporiasis" = getTable1Stats("chronicisosporiasisdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  #"Diagnosis: Disseminated non-TB Mycobacterial Infection" = getTable1Stats("mycobacdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  #"Diagnosis: Disseminated Mycosis" = getTable1Stats("mycosisdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Extrapulmonary Cryptococcosis/Meningitis" = getTable1Stats("cryptococcdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),   
  "Diagnosis: Cryptosporidiosis" = getTable1Stats("cryptospordiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "Diagnosis: Cytomegalovirus Disease" = getTable1Stats("cytomegdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: HIV Encephalopathy" = getTable1Stats("encephaldiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Candidiasis (Oesophageal, Bronchi, Trachea, or Lungs)" = getTable1Stats("esocandiddiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Chronic Herpes Simplex" = getTable1Stats("herpessvdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Pneumocystis Pneumonia" = getTable1Stats("pcpdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),   
  "Diagnosis: Progressive Multifocal Leukoencephalopathy" = getTable1Stats("pmldiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  
  # TB Prophylaxis
  "anti-TB med: Rifabutin" = getTable1Stats("rifabutin","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "anti-TB med: Rifafour(RHZE)" = getTable1Stats("rifafour","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),       
  "anti-TB med: Rifampicin(R)" = getTable1Stats("rifampicin","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),   
  "anti-TB med: Rifater(RHZ)" = getTable1Stats("rifater","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "anti-TB med: Rifinah(RH)" = getTable1Stats("rifinah","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "anti-TB med: Streptomycin" = getTable1Stats("streptomycin","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "anti-TB med: 3-FDC(RHE)" = getTable1Stats("three_FDC","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "anti-TB med: Ethambutol(E)" = getTable1Stats("ethambutol","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "anti-TB med: Ethambutol(E)" = getTable1Stats("ethambutolinh","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "anti-TB med: INH" = getTable1Stats("inhtbtx","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "anti-TB med: Other" = getTable1Stats("otherTbTx","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),       
   #"anti-TB med: Multi-Drug Resistant Tx" = getTable1Stats("mdrTbTx","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "anti-TB med: Pyrazinamide(Z)" = getTable1Stats("pyrazinamide","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "TB Prophylaxis: INH" = getTable1Stats("inh","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  
  
  "OI Prophylaxis" = getTable1Stats("oiProph","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "TB treatment" = getTable1Stats("tbtx","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  
 
  
   # ARVS
  "# current ARVs" = getTable1Stats("numARVs","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "ARVs taken last 7 days" = getTable1Stats("PillsTakenLast7Days","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "ART adherence: percent taken since last visit" = getTable1Stats("art_adherence","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),   
  "ARV Adherence- combines formats from both the old and the new CTC 2" = getTable1Stats("art_adherence2","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "ARV adherence since last visit" = getTable1Stats("arvadherence","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var)
  #"ART regimen" = getTable1Stats("regimen","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  
 
 
          
          )%>%
   htmlTable( caption  = "<b> Clinical Profile Prior to ART Initiation | Viral Load Failure 1 </b>",useViewer=T,ctable = TRUE, align = 'lcccc',
              n.cgroup = c(1,2,1 ), cgroup = c('', 'First Virologic Failure', '') ,tfoot="<sup>&Dagger;</sup>"
              )%>% htmlTable::addHtmlTableStyle(col.rgroup = c("#FFF", "#EEF")) 



```


### Clinical Profile Prior to VL 1 Date | Viral Load Failure 1
```{r fig.align='center',warning=FALSE, cache=T}

refdate='vl1_date'
              
# Generate baseline clinical covariates
Cough.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'Cough', refdate)%>%select(ptidno, Cough=var_new)
KSdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'KSdiag', refdate)%>%select(ptidno, KSdiag=var_new)
NephroCardiomyopathy.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'NephroCardiomyopathy', refdate)%>%select(ptidno, NephroCardiomyopathy=var_new)
PillsTakenLast7Days.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'PillsTakenLast7Days', refdate)%>%select(ptidno, PillsTakenLast7Days=var_new)
TBsignNone.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'TBsignNone', refdate)%>%select(ptidno, TBsignNone=var_new)
WHOstage.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'WHOstage', refdate)%>%select(ptidno, WHOstage=var_new)

cd4.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'cd4', refdate)%>%select(ptidno, cd4=var_new)
cervicalcadiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'cervicalcadiag', refdate)%>%select(ptidno, cervicalcadiag=var_new)
chronicisosporiasisdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'chronicisosporiasisdiag', refdate)%>%select(ptidno, chronicisosporiasisdiag=var_new)
coughTB.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'coughTB', refdate)%>%select(ptidno, coughTB=var_new)

cryptococcdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'cryptococcdiag', refdate)%>%select(ptidno, cryptococcdiag=var_new)
cryptospordiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'cryptospordiag', refdate)%>%select(ptidno, cryptospordiag=var_new)
cytomegdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'cytomegdiag', refdate)%>%select(ptidno, cytomegdiag=var_new)
encephaldiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'encephaldiag', refdate)%>%select(ptidno, encephaldiag=var_new)
esocandiddiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'esocandiddiag', refdate)%>%select(ptidno, esocandiddiag=var_new)
ethambutol.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'ethambutol', refdate)%>%select(ptidno, ethambutol=var_new)
ethambutolinh.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'ethambutolinh', refdate)%>%select(ptidno, ethambutolinh=var_new)
herpessvdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'herpessvdiag', refdate)%>%select(ptidno, herpessvdiag=var_new)

#// new
Weight.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'Weight', refdate)%>%select(ptidno, weight=var_new)
WeightLoss.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'WeightLoss', refdate)%>%select(ptidno, WeightLoss=var_new)
art_adherence.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'art_adherence', refdate)%>%select(ptidno, art_adherence=var_new)
art_adherence2.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'art_adherence2', refdate)%>%select(ptidno, art_adherence2=var_new)
arvadherence.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'arvadherence', refdate)%>%select(ptidno, arvadherence=var_new)
bmi.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'bmi', refdate)%>%select(ptidno, BMI=var_new)
hgb.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'hgb', refdate)%>%select(ptidno, hgb=var_new)
inh.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'inh', refdate)%>%select(ptidno, inh=var_new)
inhtbtx.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'inhtbtx', refdate)%>%select(ptidno, inhtbtx=var_new)
lymphomadiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'lymphomadiag', refdate)%>%select(ptidno, lymphomadiag=var_new)
mdrTbTx.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'mdrTbTx', refdate)%>%select(ptidno, mdrTbTx=var_new)
mycobacdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'mycobacdiag', refdate)%>%select(ptidno, mycobacdiag=var_new)
mycosisdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'mycosisdiag', refdate)%>%select(ptidno, mycosisdiag=var_new)
nightsweats.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'nightsweats', refdate)%>%select(ptidno, nightsweats=var_new)
nightsweatsTB.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'nightsweatsTB', refdate)%>%select(ptidno, nightsweatsTB=var_new)

numARVs.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'numARVs', refdate)%>%select(ptidno, numARVs=var_new)
oiProph.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'oiProph', refdate)%>%select(ptidno, oiProph=var_new)
onARV.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'onARV', refdate)%>%select(ptidno, on_ARV=var_new)
otherTbTx.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'otherTbTx', refdate)%>%select(ptidno, otherTbTx=var_new)
pcpdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'pcpdiag', refdate)%>%select(ptidno, pcpdiag=var_new)
pmldiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'pmldiag', refdate)%>%select(ptidno, pmldiag=var_new)
pyrazinamide.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'pyrazinamide', refdate)%>%select(ptidno, pyrazinamide=var_new)

regimen.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'regimen', refdate)%>%select(ptidno, regimen=var_new)
rifabutin.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rifabutin', refdate)%>%select(ptidno, rifabutin=var_new)
rifafour.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rifafour', refdate)%>%select(ptidno, rifafour=var_new)
rifampicin.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rifampicin', refdate)%>%select(ptidno, rifampicin=var_new)
rifater.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rifater', refdate)%>%select(ptidno, rifater=var_new)
rifinah.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rifinah', refdate)%>%select(ptidno, rifinah=var_new)
rsbPneumoniaDiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'rsbPneumoniaDiag', refdate)%>%select(ptidno, rsbPneumoniaDiag=var_new)
salmonelladiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'salmonelladiag', refdate)%>%select(ptidno, salmonelladiag=var_new)

streptomycin.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'streptomycin', refdate)%>%select(ptidno, streptomycin=var_new)
tbextrapulmdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'tbextrapulmdiag', refdate)%>%select(ptidno, tbextrapulmdiag=var_new)
tbpulmdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'tbpulmdiag', refdate)%>%select(ptidno, tbpulmdiag=var_new)
tbtx.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'tbtx', refdate)%>%select(ptidno, tbtx=var_new)
three_FDC.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'three_FDC', refdate)%>%select(ptidno, three_FDC=var_new)
wastingsyndiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'wastingsyndiag', refdate)%>%select(ptidno, wastingsyndiag=var_new)
toxoplasdiag.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'toxoplasdiag', refdate)%>%select(ptidno, toxoplasdiag=var_new)
weightlossTB.cx.arv=getBaselineOnArvStartWithinXMonths(patient.long,1,'weightlossTB', refdate)%>%select(ptidno, weightlossTB=var_new)
              

# add some mpee
dataset.cx.vl=dataset%>%
                   left_join(Cough.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(KSdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(NephroCardiomyopathy.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(PillsTakenLast7Days.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(TBsignNone.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(WHOstage.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(cd4.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(cervicalcadiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(chronicisosporiasisdiag.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(coughTB.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(cryptococcdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(cryptospordiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(cytomegdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(encephaldiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(esocandiddiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(ethambutol.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(ethambutolinh.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(herpessvdiag.cx.arv,  c("ptidno"="ptidno"))%>%
  
                    # new
                   left_join(Weight.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(WeightLoss.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(art_adherence.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(art_adherence2.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(arvadherence.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(bmi.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(hgb.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(inh.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(inhtbtx.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(lymphomadiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(mdrTbTx.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(mycobacdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(mycosisdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(onARV.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(nightsweats.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(nightsweatsTB.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(numARVs.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(oiProph.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(otherTbTx.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(pcpdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(pmldiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(pyrazinamide.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(regimen.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rifabutin.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rifampicin.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rifafour.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rifater.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rifinah.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(rsbPneumoniaDiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(streptomycin.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   left_join(salmonelladiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(tbextrapulmdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(tbpulmdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(tbtx.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(three_FDC.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(wastingsyndiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(toxoplasdiag.cx.arv,  c("ptidno"="ptidno"))%>%
                   left_join(weightlossTB.cx.arv,  c("ptidno"="ptidno"))%>%
  
                   mutate(
                      civilstatusENROL=ifelse(program=="AMPATH",civilstatus,civilstatusENROL)
                     
                    )%>%
                   mutate(across(binary.var, as.character))%>%
                   mutate(across(binary.var, binarize))%>%
                   mutate(across(binary.var, as.factor))
  
na.rm.var=F;stats=T;dataset.cx=dataset.cx.vl;

# Getting descriptive statistics 
mergeDesc(
  
  # vitals
  "current ARV status" = getTable1Stats("on_ARV","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "weight" = getTable1Stats("weight","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "Weight Loss" = getTable1Stats("WeightLoss","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),       
  "BMI" = getTable1Stats("BMI","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "hemoglobin" = getTable1Stats("hgb","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "CD4" = getTable1Stats("cd4","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Cough" = getTable1Stats("Cough","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "Night Sweats" = getTable1Stats("nightsweats","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  
  # staging
  "WHO stage" = getTable1Stats("WHOstage","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Staging: Symptomatic HIV-associated Nephropathy/Cardiomyopathy" = getTable1Stats("NephroCardiomyopathy","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  
   # TB symptoms
  "TB sign: none" = getTable1Stats("TBsignNone","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "TB sign: cough" = getTable1Stats("coughTB","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "TB sign: weight loss" = getTable1Stats("weightlossTB","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),

  
   # Diagnosis
  "Diagnosis: recurrent severe bacterial pneumonia" = getTable1Stats("rsbPneumoniaDiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  #"Diagnosis: Recurrent Septicemia including non-typhoidal Salmonella" = getTable1Stats("salmonelladiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: HIV Wasting Syndrome" = getTable1Stats("wastingsyndiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Toxoplasmosis, CNS" = getTable1Stats("toxoplasdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "Diagnosis: Extrapulmonary TB" = getTable1Stats("tbextrapulmdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "Diagnosis: Pulmonary TB" = getTable1Stats("tbpulmdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Lymphoma (cerebral or B-cell non-Hodgkin)" = getTable1Stats("lymphomadiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),   
  "Diagnosis: Kaposi's Sarcoma" = getTable1Stats("KSdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),       
  "Diagnosis: Cervical Cancer" = getTable1Stats("cervicalcadiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  #"Diagnosis: Chronic Isosporiasis" = getTable1Stats("chronicisosporiasisdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  #"Diagnosis: Disseminated non-TB Mycobacterial Infection" = getTable1Stats("mycobacdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  #"Diagnosis: Disseminated Mycosis" = getTable1Stats("mycosisdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Extrapulmonary Cryptococcosis/Meningitis" = getTable1Stats("cryptococcdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),   
  "Diagnosis: Cryptosporidiosis" = getTable1Stats("cryptospordiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "Diagnosis: Cytomegalovirus Disease" = getTable1Stats("cytomegdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: HIV Encephalopathy" = getTable1Stats("encephaldiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Candidiasis (Oesophageal, Bronchi, Trachea, or Lungs)" = getTable1Stats("esocandiddiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Chronic Herpes Simplex" = getTable1Stats("herpessvdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "Diagnosis: Pneumocystis Pneumonia" = getTable1Stats("pcpdiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),   
  "Diagnosis: Progressive Multifocal Leukoencephalopathy" = getTable1Stats("pmldiag","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  
  # TB Prophylaxis
  "anti-TB med: Rifabutin" = getTable1Stats("rifabutin","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "anti-TB med: Rifafour(RHZE)" = getTable1Stats("rifafour","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),       
  "anti-TB med: Rifampicin(R)" = getTable1Stats("rifampicin","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),   
  "anti-TB med: Rifater(RHZ)" = getTable1Stats("rifater","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "anti-TB med: Rifinah(RH)" = getTable1Stats("rifinah","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "anti-TB med: Streptomycin" = getTable1Stats("streptomycin","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "anti-TB med: 3-FDC(RHE)" = getTable1Stats("three_FDC","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "anti-TB med: Ethambutol(E)" = getTable1Stats("ethambutol","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "anti-TB med: Ethambutol(E)" = getTable1Stats("ethambutolinh","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  "anti-TB med: INH" = getTable1Stats("inhtbtx","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "anti-TB med: Other" = getTable1Stats("otherTbTx","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),       
   #"anti-TB med: Multi-Drug Resistant Tx" = getTable1Stats("mdrTbTx","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "anti-TB med: Pyrazinamide(Z)" = getTable1Stats("pyrazinamide","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "TB Prophylaxis: INH" = getTable1Stats("inh","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  
  
  "OI Prophylaxis" = getTable1Stats("oiProph","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  "TB treatment" = getTable1Stats("tbtx","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  
 
  
   # ARVS
  "# current ARVs" = getTable1Stats("numARVs","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),  
  #"ARVs taken last 7 days" = getTable1Stats("PillsTakenLast7Days","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "ART adherence: percent taken since last visit" = getTable1Stats("art_adherence","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),   
  "ARV Adherence- combines formats from both the old and the new CTC 2" = getTable1Stats("art_adherence2","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var), 
  "ARV adherence since last visit" = getTable1Stats("arvadherence","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var)
  #"ART regimen" = getTable1Stats("regimen","vl_failure1",desc_both,dataset.cx,statistics=stats,na.rm=na.rm.var),
  
 
 
          
          )%>%
   htmlTable( caption  = "<b> Clinical Profile Prior to VL 1 Measurement | Viral Load Failure 1 </b>",useViewer=T,ctable = TRUE, align = 'lcccc',
              n.cgroup = c(1,2,1 ), cgroup = c('', 'First Virologic Failure', '') ,tfoot="<sup>&Dagger;</sup>"
              )%>% htmlTable::addHtmlTableStyle(col.rgroup = c("#FFF", "#EEF")) 



```








```{r  results='asis', fig.keep='all', echo=FALSE, results='asis', fig.keep='all', eval=F }

idata=dataset %>%
 #filter(program=="AMPATH")%>%
  mutate(year = as.numeric(format(vl1_date,'%Y')), on_ART=factor(ifelse(arvstartTHERAP<=vl1_date, "Yes", "No")))%>% 
  filter(!is.na(vl_count1))

xts( idata$vl_count1, order.by=idata$vl1_date ) %>% 
  .['2014/2020-12-31']  %>% 
  apply.monthly(. , mean) %>% 
  ts( ., start = c(2014,1,1), frequency = 12) %>% 
  ts_decompose(., type = "all") 
  #ts_heatmap(.)

```





```{r fig.align='center',warning=FALSE, cache=T, eval=F}

idata=dataset %>%
 #filter(program=="MBARARA")%>%
  mutate(year = as.numeric(format(vl1_date,'%Y')), on_ART=factor(ifelse(arvstartTHERAP<=vl1_date, "Yes", "No")))%>% 
  filter(!is.na(num_days_to_vl1))

xts( idata$num_days_to_vl1, order.by=idata$vl1_date ) %>% 
  .['2014/2020-12-31']  %>% 
  apply.monthly(. , mean) %>% 
  ts( ., start = c(2014,1,1), frequency = 12) %>% 
  ts_seasonal(., type = 'all', title = "Trends")
 # ts_heatmap(.)

```


## Bivariate Plots

To understand the relationship of each continuous variable with the y variable:

1. Plot box plot for each of the variables to do a visual comparison between the groups
2. Plot the explanatory variable distribution for both the variables to understand the variability uniquely explained (The non-intersecting part of the blue and the red is the variation explained by the variable)

## Socio-demographics

```{r fig.align='center', cache=T, fig.width=6, warning=FALSE}
y <- c("vl_failure2","vl_failure1")
x <- c('male','Death','civilstatusENROL', 'transfer','Height','Weight', 'entrypoint', 'nhif', 'bmi','program', "ageatarvstart", "ageatarvstartTHERAP",
       'num_days_to_vl1','num_days_to_vl2','num_days_to_vl3', 'vl_count1','vl_count2','vl_count3', 'num_encs_to_vl1','num_encs_to_vl2','num_encs_to_vl3'
       )
raw_df=dataset%>%select(c(y,x))
cont_univ_df <- raw_df %>% select_if(is.numeric) %>% dplyr::mutate(row_no =1:n() )

for (value in  colnames(cont_univ_df[-ncol(cont_univ_df)]))  { 
  print(value)
  p1 <- ggplot2.histogram(raw_df%>%drop_na(vl_failure1), xName=value, backgroundColor="white",  bins=10,
                         #brewerPalette="Paired",
                         groupName="vl_failure1", legendPosition="top", xScale="log2",
                         
                         alpha=0.5, addDensity=TRUE,  removePanelGrid=TRUE ,removePanelBorder=TRUE,
                         addMeanLine=TRUE, meanLineSize=1)+ theme_minimal()+ theme(legend.position="top")
   p2<- ggboxplot(raw_df%>%drop_na(vl_failure1),   alpha=0.2, x = "vl_failure1", y = value, color = "vl_failure1", palette = "jco", add = "jitter")+
          stat_compare_means()+ guides(fill=FALSE, color=FALSE)	+ theme_minimal()
   
  
   p3 <- ggplot2.histogram(raw_df%>%drop_na(vl_failure2), xName=value, backgroundColor="white",  bins=10,
                         #brewerPalette="Paired",
                         groupName="vl_failure2", legendPosition="top", xScale="log2",
                         
                         alpha=0.5, addDensity=TRUE,  removePanelGrid=TRUE ,removePanelBorder=TRUE,
                         addMeanLine=TRUE, meanLineSize=1)+ theme_minimal()+ theme(legend.position="top")
   p4<- ggboxplot(raw_df%>%drop_na(vl_failure2), alpha=0.2, x = "vl_failure2", y = value, color = "vl_failure2", palette = "jco", add = "jitter")+
    stat_compare_means()+ guides(fill=FALSE, color=FALSE)	+ theme_minimal()
   
   grid.arrange( p1,p2, p3,p4,nrow=2) 
  # print(p2 )
}


#names(dataset)

```



# Univariate Analysis

For each continuous column, visually check the following: 

1. Variation in the column
2. Its distribution
3. Any outliers
4. q-q plot with normal distribution


```{r fig.align='center', cache=T, fig.width=5, warning=FALSE}


for(column in colnames(cont_univ_df[-ncol(cont_univ_df)])){
  
  print(column)
  p1 <- ggplot(cont_univ_df, aes_string(x='row_no', y= column)) +
    geom_point(show.legend = FALSE) +
    labs(x = 'Univariate plot', y=column) +
    ggtitle(column) +
    theme_minimal()

  # Cumulative plot
  legendcols <- c("Normal distribution"="darkred","Density"="darkBlue","Histogram"="lightBlue")
  p2 <- ggplot(cont_univ_df,aes_string(x = column)) +
    geom_histogram(aes(y=..density.., fill ="Histogram"), bins = 50) +
    stat_function(fun = dnorm, aes(color="Normal distribution"),  size = 1,
                args = list(mean = mean(cont_univ_df[[column]]),
                            sd = sd(cont_univ_df[[column]]))) +
  geom_density(aes(y=..density.., color="Density"),  size = 1)+
  scale_colour_manual(name="Distribution",values=legendcols) +
  scale_fill_manual(name="Bar",values=legendcols) +
  theme_minimal() + theme(legend.position="none")

  p3 <- ggplot(cont_univ_df %>% mutate(constant = column),
    aes_string(x="constant", y= column, group = 123)) +
    geom_boxplot() +
    labs(y=column) +
    theme_minimal()

  p4 <- ggplot(cont_univ_df,aes_string(sample = column)) +
    stat_qq() + stat_qq_line() +
    theme_minimal()
  grid.arrange(p1, p2, p3, p4, nrow=2)

}
```


For categorical variables,visually check the frequencies and percentages.

```{r fig.align='center',cache=T,  fig.width=5, warning=FALSE}

univ_cat_df <- raw_df %>% select_if(function(col) {is.factor(col) | is.character(col)})
plist= list()
for(column in colnames(univ_cat_df)){
gg = ggplot(univ_cat_df,aes_string(column)) +
    geom_bar() + coord_flip() +
    ggtitle(column) +
    theme_minimal()
plist[[column]]=gg
if(length(plist)==4){
  print(names(plist))
  do.call("grid.arrange", c(plist,  nrow=2))
  plist= list()
}
}


```

